---
title: "HDAT9600 Final Assignment"
author: "insert your team name here"
date: "insert date of completion here"
output:
  html_document: default
  pdf_document: default
subtitle: 'Submission deadline is 5pm Friday 23 2019 '
---

```{r setup, include=FALSE}
# leave this code here, but feel free to adjust the options or add some more
# see the knitr documentation for details
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=12)
```

## Instructions

This file (hdat9600_final_assignment.Rmd) is the R Markdown document in which you need to complete your HDAT9600 final assignment. This assignment is assessed and will count for 30% of the total course marks. The assignment comprises two tasks worth 15 marks each. The first task will focus on logistic regression, and the second task will focus on survival analysis. There is no word limit, but a report of about 10 pages in length when printed (except that it will not be printed) is appropriate.

Don't hesitate to ask the course convenor for help via OpenLearning. The course instructor are happy to point you in the right direction and to make suggestions, but they won't, of course, complete your assignments for you!


## Data for this assignment

The data used for this assignment consist of records from Intensive Care Unit (ICU) hospital stays in the USA. All patients were adults who were admitted for a wide variety of reasons. ICU stays of less than 48 hours have been excluded. 

The source data for the assignment are data made freely available for the 2012 MIT PhysioNet/Computing for Cardiology Challenge. Details are provided [here](https://physionet.org/challenge/2012/). Training Set A data have been used. The original data has been modified and assembled to suit the purpose of this assignment. Full details of the preparatory work can be found in the hdat9600-final-assignment-data-preparation file.

The dataframe consists of 120 variables, which are defined as follows:

####Patient Descriptor Variables
<li> <em>RecordID:</em>   a unique integer for each ICU stay</li>
<li> <em>Age:</em>                 years</li>
<li> <em>Gender:</em>              male/female</li>
<li> <em>Height:</em>              cm</li>
<li> <em>ICUType:</em>             Coronary Care Unit; Cardiac Surgery Recovery Unit; Medical ICU; Surgical ICU</li>
<li> <em>Length_of_stay:</em>      The number of days between the patient’s admission to the ICU and the end of hospitalisation</li>
<li> <em>Survival:</em>            The number of days between ICU admission and death for patients who died</li>


####Outcome Variables
<li> <em>in_hospital_death:</em>   0:survivor/1:died in-hospital **this is the outcome variable for Task 1: Logistic Regression**</li>
<li> <em>Status:</em>              True/False **this is the censoring variable for Task 2: Survival Analysis**</li>
<li> <em>Days:</em>                Length of survival (in days) **this is the survival time variable for Task 2: Survival Analysis**</li>


####Clinical Variables
<em>Use the hyperlinks below to find out more about the clinical meaning of each variable.</em>
The first two clinical variables are summary scores that are used to assess patient condition and risk.

<li> <em>SAPS-I score</em> [Simplified Acute Physiological Score (<a href="http://www.ncbi.nlm.nih.gov/pubmed/6499483"
target="_blank"><em>Le Gall et al., 1984</em></a>)]</li>
<li> <em>SOFA score</em> [Sequential Organ Failure Assessment (<a href="http://www.ncbi.nlm.nih.gov/pubmed/11594901"
target="_blank"><em>Ferreira et al., 2001</em></a>)]</li>

###

The following 36 clinical measures were assessed at multiple timepoints during each patient's ICU stay. For each of the 36 clinical measures, you are given 3 summary variables: a) The minimum value during the first 24 hours in ICU (_min), b) The maximum value during the first 24 hours in ICU (_max), and c) The difference between the mean and the most extreme values during the first 24 hours in ICU (_diff). For example, for the clinical measure Cholesterol, these three variables are labelled 'Cholesterol_min', 'Cholesterol_max', and 'Cholesterol_diff'.

<li><a href="http://en.wikipedia.org/wiki/Human_serum_albumin" target="_blank">
<em>Albumin</em></a> (g/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Alkaline_phosphatase" target="_blank">
<em>ALP</em></a> [Alkaline phosphatase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Alanine_transaminase" target="_blank">
<em>ALT</em></a> [Alanine transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Aspartate_transaminase" target="_blank">
<em>AST</em></a> [Aspartate transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bilirubin" target="_blank">
<em>Bilirubin</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/BUN" target="_blank">
<em>BUN</em></a> [Blood urea nitrogen (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Cholesterol" target="_blank">
<em>Cholesterol</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_creatinine#Plasma_creatinine"
target="_blank">
<em>Creatinine</em></a> [Serum creatinine (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>DiasABP</em></a> [Invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/FIO2" target="_blank">
<em>FiO2</em></a> [Fractional inspired O<sub>2</sub> (0-1)]</li>
<li><a href="http://en.wikipedia.org/wiki/Glasgow_coma_score" target="_blank">
<em>GCS</em></a> [Glasgow Coma Score (3-15)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_glucose" target="_blank">
<em>Glucose</em></a> [Serum glucose (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bicarbonate#Diagnostics"
target="_blank">
<em>HCO3</em></a> [Serum bicarbonate (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hematocrit" target="_blank">
<em>HCT</em></a> [Hematocrit (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Heart_rate" target="_blank">
<em>HR</em></a> [Heart rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hypokalemia" target="_blank">
<em>K</em></a> [Serum potassium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Lactic_acid" target="_blank">
<em>Lactate</em></a> (mmol/L)</li>
<li><a href="http://en.wikipedia.org/wiki/Magnesium#Biological_role"
target="_blank">
<em>Mg</em></a> [Serum magnesium (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>MAP</em></a> [Invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mechanical_ventilation"
target="_blank">
<em>MechVent</em></a> [Mechanical ventilation respiration (0:false, or 1:true)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_sodium" target="_blank">
<em>Na</em></a> [Serum sodium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>NIDiasABP</em></a> [Non-invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>NIMAP</em></a> [Non-invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>NISysABP</em></a> [Non-invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaCO2</em></a> [partial pressure of arterial CO<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaO2</em></a> [Partial pressure of arterial O<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>pH</em></a> [Arterial pH (0-14)]</li>
<li><a href="http://en.wikipedia.org/wiki/Platelets" target="_blank">
<em>Platelets</em></a> (cells/nL)</li>
<li><a href="http://en.wikipedia.org/wiki/Respiratory_physiology" target="_blank">
<em>RespRate</em></a> [Respiration rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>SaO2</em></a> [O<sub>2</sub> saturation in hemoglobin (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>SysABP</em></a> [Invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Normal_human_body_temperature"
target="_blank">
<em>Temp</em></a> [Temperature (&deg;C)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropI</em></a> [Troponin-I (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropT</em></a> [Troponin-T (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Fluid_balance" target="_blank">
<em>Urine</em></a> [Urine output (mL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Reference_ranges_for_blood_tests#Hematology" target="_blank">
<em>WBC</em></a> [White blood cell count (cells/nL)]</li>
<li>
<em>Weight</em> (kg)</li>


##Accessing the Data
The data frame can be loaded with the following code:

```{r}
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
```

# Task 1 (15 marks)

In this task, you are required to develop a logistic regression model using the `icu_patients_df1` data set which adequately explains or predicts the `in_hospital_death` variable as the outcome using a subset of the available predictor variables. You should fit a series of models, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge. It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct or complete, but do you best). Aim for between five and ten predictor variables (slightly more or fewer is OK). You should assess each model you consider for goodness of fit and other relevant statistics to help you choose between them. For your final model, present a set of diagnostic statistics and/or charts and comment on them. You don't need to do an exhaustive exploratory data analysis of all the variables in the data set, but you should examine those variables that you use in your model. Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find compared to the same model fitted to the imputed data. 


### Hints

1. Select an initial subset of explanatory variables that you will use to predict the risk of in-hospital death. Justify your choice.

2. Conduct basic exploratory data analysis on your variables of choice.

3. Fit appropriate univariate logistic regression models.

4. Fit an appropriate series of multivariable logistic regression models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

5. Present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge.

6. For your final model, present a set of diagnostic statistics and/or charts and comment on them.

7. Write a paragraph summarising the most important findings of your final model. Include the most important values from the statistical output, and a simple clinical interpretation.



#####
**Create your response to this task here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text.** 
```{r, setup for task1}
library(tidyverse)
```
##Jason
##Consideration for variables
We only need to include 5-10ish variables. This is requires a logical approach to culling. The first step is to remove any variables that do not have a full set of data (eg 2061 records). The rationale for this is that if you include incomplete variables then drop an incomplete variable you may not be able to adequately compare the current model to the previous one.  
  
```{r data cleaning and checking}

#check which variables have incomplete data
nalist <- sapply(icu_patients_df1, function(x) sum(is.na(x)))
nalist[nalist!=0]

#check data for internal consistency (max vs min values)
for (i in names(icu_patients_df1)){
  if (substr(i, nchar(i)-2, nchar(i)) == "max"){
  j <- paste0(substr(i, 1, nchar(i)-4 ),"_min")
  print(c(substr(i, 1, nchar(i)-4 ), length(as.matrix(icu_patients_df1[na.omit(icu_patients_df1[,i])<na.omit(icu_patients_df1[,j]),c(i,j)]))/2))
  }
}

#ensure all max and min values are in the right columns
#function to swap values that are inconsistent with allocation
maxval <- function(df, Max, Min){
  df[na.omit(df[,Max])<na.omit(df[,Min]),c(Min,Max)] <- unname(df[na.omit(df[,Max])<na.omit(df[,Min]),c(Max,Min)])
  df
}

#loop checks for "max" variables then compares to the corresponding "min" value and swaps values where necessary
for (i in names(icu_patients_df1)){
  if (substr(i, nchar(i)-2, nchar(i)) == "max"){
  j <- paste0(substr(i, 1, nchar(i)-4 ),"_min")
  icu_patients_df1 <- maxval(icu_patients_df1, i, j)
  }
}

#reassess internal consistency (max vs min values)
for (i in names(icu_patients_df1)){
  if (substr(i, nchar(i)-2, nchar(i)) == "max"){
  j <- paste0(substr(i, 1, nchar(i)-4 ),"_min")
  print(c(substr(i, 1, nchar(i)-4 ), length(as.matrix(icu_patients_df1[na.omit(icu_patients_df1[,i])<na.omit(icu_patients_df1[,j],c(i,j))]))/2))
  }
}

#check score variable distributions
table(icu_patients_df1$SOFA)
table(icu_patients_df1$SAPS1)

#65 scores of -1 for SOFA - investigate (check for consistency with scoring against GCS, MAP, Creatinine/Urine, PaO2/FiO2, Bilirubin and Platelets)
icu_patients_df1[icu_patients_df1$SOFA==0 & icu_patients_df1$MAP_min<70,c("SOFA", "PaO2_min", "FiO2_max", "Bilirubin_max", "GCS_min", "Creatinine_max", "Urine_max", "Platelets_min", "MAP_min")]
#noted variation in score relative to variables. Eg, 20/52 SOFA scores of 0 appear wrong based on MAP. Difficult and time consuming to determine accuracy of other scores - assume inaccurate, leave out. Use SAPS1 instead

#remove incomplete variables to maintain data integrity but keeps SAPS1 as a predictor and not too many values missing (96/2061)
ICUMort <- icu_patients_df1[,complete.cases(t(icu_patients_df1))]
ICUMort <- cbind(ICUMort, SAPS1=icu_patients_df1$SAPS1)
summary(ICUMort)
str(ICUMort)

#SAPS1 includes Age, HR, SBP, Temp, RR, urine, BUN, HCT, WBC, glucose, potassium, sodium, HCO3, GCS using most deranged value in first 24 hours.

```
  
#Variables left after incomplete and other outcome related variables removed.

##Assessable outcome: "in_hospital_death"  
  
_Demographics: "Age", "Gender", "ICUType"_
_Age and sex are demographics important for identifying confounders. Height is missing a lot of values. Is length of stay a predictor or an outcome? Suggest outcome, remove.  Weight missing a lot of values. ICUType identifies different patient cohorts._
##Selected Variables: "Age", "Gender", "ICUType" 
Total number of variables = 3

_Scores: "SOFA", "SAPS1"_
_SOFA score uses GCS, MAP, Creatinine/Urine, PaO2/FiO2, Bilirubin and Platelets. Upon closer inspect of SOFA score, it appears the calculations may be wildly inaccurate. Don't include. SAPS1 includes Age, HR, SBP, Temp, RR, urine, BUN, HCT, WBC, glucose, potassium, sodium, HCO3, GCS using most deranged value in first 24 hours._
##Selected Variables: "SAPS1"
Total number of variables = 4

_CNS: "GCS", "Na"_
_Sodium and GCS included in SAPS1. Review both in context of other variables._
##Selected Variables: "Na", "GCS"
Total number of variables = 6

_Cardiovascular: HR", "K", "MAP", "Mg", "TroponinI", "TroponinT"_
_TroponinI is the value that is routinely checked for patient's who present with chest pain/MI. The dataset has a coronary care unit and a cadiac surgical ICU, include TnI(not TnT). Temp is bad if high or low, can be a marker for infection. Likely WBC will be better. HR bad if too high or too low, marker for SIRS. Unsure about usefulness of including electrolytes (K, Mg) as they are quite modifiable, leave out._
##Selected Variables: "MAP", "TroponinI"
Total number of variables = 8

_Respiratory: "FiO2", "PaCO2", "PaO2", "RespRate", "SaO2"_
_PaCO2 can be bad low and high but is also easily modified, leave out. Respiratory rate bad if low or high but likely to be worse in ICU if high. SOFA score no good, use PaO2. Lower is worse._
##Selected Variables: "RespRate", "PaO2"
Total number of variables = 10

_Renal: "BUN", "Creatinine", "Urine"_
_All values are good markers of acute kidney injury or chronic kidney disease which both have a higher than average mortality rate._
##Selected Variables: "BUN"
Total number of variables = 11

_Liver: "Albumin", "ALP", "ALT", "AST", "Bilirubin", "Cholesterol"_
_Low albumin is bad. High bilirubin is bad. ALT and ALT are supportive markers for liver failure but this is adequately covered by bilirubin. Unclear about the role of Cholesterol in critical illness, leave out._
##Selected Variables: "Bilirubin"
Total number of variables = 12

_Metabolic/Other: "Glucose", "HCO3", "HCT", "Lactate", "pH", "Platelets", "Temp", "WBC"_
_Glucose bad if low or high, diabetes also a high mortaility comorbidity in ICU. HCO3 bad if it’s low. High lactate is bad. WBC marker for infection in both directions. Low platelets are bad. pH is a good marker of severity of illness, bad in both directions, also related to HCO3 and PaCO2. HCT can be related to life threatening disorders._
##Selected Variables: "Glucose", "WBC", "pH", "Platelets"
Total number of variables = 16

#16 Variables selected with intention to drop ~5+.

```{r review of selected variables}
#create dataframe with selected variables
#create list of selected variables
var.list <- c("Age", "Gender", "ICUType", "SAPS1", "Na", "GCS", "MAP", "TroponinI", "RespRate", "PaO2", "BUN", "Bilirubin", "Glucose", "WBC", "pH", "Platelets", "in_hospital_death")

#create empty list for variables to go into
test.list <- NULL

#fill empty list with all possible variables for data frame
for (i in names(ICUMort)){
  for (j in var.list){
    if (substr(i, 1, nchar(j)) == j){
      test.list <- c(test.list, i) 
      }
    }
  }

#check list
test.list

#update to testing data frame
ICUMort.t <- ICUMort[,test.list]
#remove na data from data frame for consistent analysis
ICUMort.t <- ICUMort.t[complete.cases(ICUMort.t),]

#put all variables in a model to determine statistical significance
testing_mod <- glm(in_hospital_death ~ ., data=ICUMort.t, family="binomial")
(sumtest <- summary(testing_mod))

#print statisically significant values (at  0.1) to assess
coef.df <- as.data.frame(sumtest$coefficients)
round(as.matrix(coef.df[coef.df$`Pr(>|z|)`<0.1,]),4)

#statisically significant values -> Age, Bilirubin_diff, BUN_min, GCS_max, ICUType, MAP_max, Na_max, PaO2_max, Platelets_diff, RespRate_min, SAPS1
#Gender not statistically significant alone but likely to have interactions.

#final data frame for analysis
ICUMort.f <- icu_patients_df1[,c("SAPS1", "Age", "Bilirubin_diff", "BUN_min", "GCS_max", "ICUType", "MAP_max", "PaO2_max", "Platelets_diff", "RespRate_min", "WBC_max", "Gender", "in_hospital_death")]
ICUMort.f <- ICUMort.f[complete.cases(ICUMort.f),]

```

##EDA
```{r review of selected variables}
summary(ICUMort.f)
str(ICUMort.f)

#density plots for reviewing variables based on outcome
for (i in names(ICUMort.f)){
  
  if(is.numeric(ICUMort.f[[i]])){
  
    print(ggplot(ICUMort.f, aes(ICUMort.f[[i]])) +
          geom_histogram(aes(y = ..density.., fill = factor(in_hospital_death, labels = c("No","Yes"))), alpha = .3, bins = 60, position = "identity") +
          labs(x=i, fill="Died", title = paste0("Density Plot of ", i,", ", length(na.omit(ICUMort.f[[i]]))," values available")))

    }else{

    print(ggplot(ICUMort.f, aes(ICUMort.f[[i]], fill=factor(in_hospital_death, labels = c("No", "Yes")))) +
          geom_bar(position = "fill") +
          labs(x = i, fill = "Died", title = paste0("Bar Plot of ", i,", ",  length(na.omit(ICUMort.f[[i]])), " values available")))
 
  }
}

#review relationships with gender 
for (i in names(ICUMort.f[,names(ICUMort.f)!="in_hospital_death"])){
      
        if(is.numeric(ICUMort.f[[i]])){
        
          ybl <- boxplot.stats(ICUMort.f[[i]])$stats[c(1, 5)]
          ybl <- ybl * c(.75,1.25)
          print(ggplot(ICUMort.f, aes(Gender,ICUMort.f[[i]], fill = factor(in_hospital_death, labels = c("No","Yes")))) +
                geom_boxplot(outlier.colour = NA) + coord_cartesian(ylim = ybl) +
                labs(y=i, fill="Died", title = paste0("Density Plot of ", i,", ", length(na.omit(ICUMort.f[[i]]))," values available"))) 
      
          }
}
#possible interactions: Glucose:Gender, BUN:Gender, RespRate:Gender, Age:Gender

#review relationships with ICUtype
for (i in names(ICUMort.f[,names(ICUMort.f)!="in_hospital_death"])){
      
        if(is.numeric(ICUMort.f[[i]])){
        
          ybl <- boxplot.stats(ICUMort.f[[i]])$stats[c(1, 5)]
          ybl <- ybl * c(.75,1.25)
          print(ggplot(ICUMort.f, aes(ICUType,ICUMort.f[[i]], fill = factor(in_hospital_death, labels = c("No","Yes")))) +
                geom_boxplot(outlier.colour = NA) + coord_cartesian(ylim = ybl) +
                labs(y=i, fill="Died", title = paste0("Density Plot of ", i,", ", length(na.omit(ICUMort.f[[i]]))," values available"))) 
      
          }
}
#possible interactions: all variables appear to have some interaction with ICU type. Explored further later.


#correlation plot to  explore relationships
library(corrplot)
ICUMort.cor <- cor(ICUMort.f[,sapply(ICUMort.f, is.numeric)])
corrplot(ICUMort.cor,  method = "circle", type = "upper")

#further explore noted correlations
#unexpected correlation
ggplot(ICUMort.f, aes(RespRate_min, GCS_max, colour = factor(in_hospital_death, labels = c("No","Yes")))) + geom_point(position = "jitter") + geom_smooth(method="lm") + labs(colour="Died")

#PaO2 not part of SAPS1, unexpected correlation
ggplot(ICUMort.f, aes(PaO2_max, SAPS1, colour = factor(in_hospital_death, labels = c("No","Yes")))) + geom_point(position = "jitter") + geom_smooth(method="lm") + labs(colour="Died")

#GCS part of SAPS1 score - expected correlation
ggplot(ICUMort.f, aes(GCS_max, SAPS1, colour = factor(in_hospital_death, labels = c("No","Yes")))) + geom_point(position = "jitter") + geom_smooth(method="lm") + labs(colour="Died")

```


```{r, univariate models}

#Should start with univariate models for each of the variables we included.
#Create vector of variables
summary(glm(in_hospital_death ~ . , family = binomial(), data = ICUMort.f))
var_selection <- c("SAPS1", "Age", "Bilirubin_diff", "BUN_min", "GCS_max", "ICUType", "MAP_max", "PaO2_max", "Platelets_diff", "RespRate_min", "WBC_max", "Gender", "in_hospital_death")
#Create a list of univariate models for each of the variables selected
formulas<- paste('in_hospital_death ~', var_selection)
unimod_list <- lapply(formulas, glm, family = binomial(), data = ICUMort.f)
#Summary of all models
unimod_summ <- lapply(unimod_list, summary)
unimod_summ
#All variables except Na_max and sex are significant in univariate models
#Extract exp of coefficients - odds for each variable
unimod_odds <- lapply(unimod_list, function(x) {exp(coef(x))})
unimod_odds

#Check interactions from EDA
var_int_selection <- c("Age:Gender", "BUN_min:Age", "GCS_max:RespRate_min", "GCS_max:Gender", "Platelets_diff:Gender", "RespRate_min:Gender", "SAPS1:ICUType", "Age:ICUType", "Bilirubin_diff:ICUType", "BUN_min:ICUType", "GCS_max:ICUType", "MAP_max:ICUType", "PaO2_max:ICUType", "Platelets_diff:ICUType", "RespRate_min:ICUType", "Gender:ICUType")
formulas_int<- paste('in_hospital_death ~', var_int_selection)
unimod_int_list <- lapply(formulas_int, glm, family = binomial(), data = ICUMort.f)
unimod_int_summ <- lapply(unimod_int_list, summary)
unimod_int_summ

#all interactions statistically significant

```


```{r, multivariate models}
#Only include statisically significant variables from above. cut off 0.05.
full_mod<-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max + 
                Platelets_diff + RespRate_min, family = binomial(), data = ICUMort.f)
summary(full_mod)
#check statistically significant values
#create statistically significant model
sig_full_mod<-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max 
                , family = binomial(), data = ICUMort.f)
summary(sig_full_mod)

#compare full with statistically significant for anova and AIC
anova(full_mod, sig_full_mod, test = "Chisq")
full_mod$aic
sig_full_mod$aic
#test not significant, AIC lower for sig model, we can accept the simpler model

#check full models against stepwise removal of variables
red_full_mod <- step(full_mod, trace = 0)
anova(red_full_mod, sig_full_mod, test = "Chisq")
red_full_mod$aic
sig_full_mod$aic
#test not significant, AIC similar, we can accept the simpler model - sig_full_mod
f.mod1 <- sig_full_mod

#Statistically significant interactions from individual models above
int_mod<-glm(in_hospital_death ~  Age:Gender + BUN_min:Age + GCS_max:RespRate_min + GCS_max:Gender + Platelets_diff:Gender + RespRate_min:Gender + SAPS1:ICUType + Age:ICUType + Bilirubin_diff:ICUType + BUN_min:ICUType + GCS_max:ICUType + MAP_max:ICUType + PaO2_max:ICUType + Platelets_diff:ICUType + RespRate_min:ICUType + Gender:ICUType, family = binomial(), data = ICUMort.f)
summary(int_mod)
#check statistically significant values
# coef.df <- as.data.frame(summary(int_mod)$coefficients)
# as.matrix(coef.df[coef.df$`Pr(>|z|)`<0.05,])
#Age:Gender, BUN_min:Gender, pH_diff:Gender, BUN_min:ICUType, PaO2_max:ICUType, GCS_max:ICUType
#create statistically significant model
sig_int_mod<-glm(in_hospital_death ~ BUN_min:Age + GCS_max:Gender + SAPS1:ICUType + Bilirubin_diff:ICUType + BUN_min:ICUType + GCS_max:ICUType + PaO2_max:ICUType, family = binomial(), data = ICUMort.f)
summary(sig_int_mod)

#compare full with statistically significant for anova and AIC
anova(int_mod, sig_int_mod, test = "Chisq")
int_mod$aic
sig_int_mod$aic
#test statistically significant, AIC lower for int model, reject null hypothesis and keep original model

#check full models against stepwise removal of variables
red_int_mod <- step(int_mod, trace = 0)
anova(red_int_mod, int_mod, test = "Chisq")
red_int_mod$aic
int_mod$aic
#test not significant, AIC better for reduced model, we can accept the simpler model - red_int_mod
f.mod2 <- red_int_mod

#Full model with interactions
full_int_mod <-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max + 
                Platelets_diff + RespRate_min + Age:Gender + BUN_min:Age + GCS_max:RespRate_min + GCS_max:Gender +
                Platelets_diff:Gender + RespRate_min:Gender + SAPS1:ICUType + Age:ICUType + Bilirubin_diff:ICUType +
                BUN_min:ICUType + GCS_max:ICUType + MAP_max:ICUType + PaO2_max:ICUType + Platelets_diff:ICUType +
                RespRate_min:ICUType + Gender:ICUType, family = binomial(), data = ICUMort.f)
summary(full_int_mod)
#check statistically significant values
# coef.df <- as.data.frame(summary(full_int_mod)$coefficients)
# as.matrix(coef.df[coef.df$`Pr(>|z|)`<0.05,])
#pH_diff, BUN_min, Na_max, GCS_max, Age, ICUType, Gender, Age:Gender, PaO2_max:ICUType, GCS_max:ICUType

#create statistically significant model at 0.1
sig_full_int_mod<-glm(in_hospital_death ~ SAPS1 + GCS_max + ICUType + PaO2_max + Age:Gender + GCS_max:Gender +
                        SAPS1:ICUType + GCS_max:ICUType,
                      family = binomial(), data = ICUMort.f)
summary(sig_full_int_mod)

#compare full with statistically significant for anova and AIC
anova(full_int_mod, sig_full_int_mod, test = "Chisq")
full_int_mod$aic
sig_full_int_mod$aic
#test statistically significant, AIC lower for full int model, reject null hypothesis and keep original model

#check full models against stepwise removal of variables
red_full_int_mod <- step(full_int_mod, trace = 0)
anova(red_full_int_mod, full_int_mod, test = "Chisq")
red_full_int_mod$aic
full_int_mod$aic
#test not significant, AIC better for reduced model, we can accept the simpler model - red_full_int_mod
f.mod3 <- red_full_int_mod

#Compare chosen models to a null model
null_mod <- glm(in_hospital_death ~ 1, family = binomial, data = ICUMort.f)
#anova(sig_full_mod, null_mod, test = "Chisq")
anova(f.mod1, null_mod, test = "Chisq")
anova(f.mod2, null_mod, test = "Chisq")
anova(f.mod3, null_mod, test = "Chisq")
#All tests statistically significant - Reject null hypothesis of no relationship between predictors and outcome for all models

#compare chosen models
anova(f.mod1, f.mod2, test = "Chisq")
f.mod1$aic
f.mod2$aic
anova(f.mod1, f.mod3, test = "Chisq")
f.mod1$aic
f.mod3$aic
anova(f.mod2, f.mod3, test = "Chisq")
f.mod2$aic
f.mod3$aic
#all tests statistically significant, can not reject null hypothesis, AIC improved on reduced models, 
#unable to compare interaction models due to not enough degrees of freedom, 
#p-value more significant for reduced full interaction model and AIC lower that both other models

```

```{r, Task1, goodness of fit}

#possible models - sig_full_mod, red_int_mod, red_full_int_mod

#Calculate the Hosmer-Lemeshow statistic
#linear predictors
library(ResourceSelection)
#Vector of predicted probabilities
pred_mod1 <- predict(f.mod1, type = "response")
(hl_mod1 <- hoslem.test(f.mod1$y, pred_mod1, g=80))
#hl_mod1
#No indication of a lack of fit as p > 0.05.

pred_mod2 <- predict(f.mod2, type = "response")
(hl_mod2 <- hoslem.test(f.mod2$y, pred_mod2, g=80))
#hl_mod2
#No indication of a lack of fit as p > 0.05.

pred_mod3 <- predict(f.mod3, type = "response")
(hl_mod3 <- hoslem.test(f.mod3$y, pred_mod3, g=80))
#lack of fit as p < 0.05.

pred_null <- predict(null_mod, type = "response")
# (hl_null <- hoslem.test(null_mod$y, pred_null, g=50))
# hl_null

#difficulty assessing goodness of fit, check p-value over bins size 11:210 bins
cn <- c("bins","f.mod1", "f.mod2", "f.mod3")
mt <- matrix(nrow=200,ncol=4, dimnames = list(NULL, cn))
for (i in 11:210){
    j<- i-10
    w <- hoslem.test(f.mod1$y, pred_mod1, g=i)
    x <- hoslem.test(f.mod2$y, pred_mod2, g=i)
    y <- hoslem.test(f.mod3$y, pred_mod3, g=i)
#    z <- hoslem.test(null_mod$y, pred_null, g=i)
      mt[j,1] <- i
      mt[j,2] <- w$p.value
      mt[j,3] <- x$p.value
      mt[j,4] <- y$p.value
#      mt[j,5] <- z$p.value
}
mt <- as.data.frame(mt)
#plot goodness of fit over 200 bins
plot(mt$bins,mt$f.mod1, type="l", ylim=c(0,1), xlab="Number of bins", ylab="p-value", main="Hosmer-Lemeshow Goodness of Fit Test", col="blue",lwd=2)
abline(h=0.05, col="red")
abline(v=80, col="red")
lines(mt$bins,mt$f.mod2, col="dark green", lty = 2,lwd=2)
lines(mt$bins,mt$f.mod3, col="black", lty = 1,lwd=2)
legend("bottomleft",legend=c(cn[-1]), fill="white", lty=c(1,2), lwd=2, col = c("blue", "dark green", "black", "orange"))

#trendline for goodness of fit
ggplot(mt, aes(bins, f.mod1)) + geom_smooth(col="blue") + geom_smooth(aes(y=f.mod2), col = "dark green", linetype = 2) + geom_smooth(aes(y=f.mod3), col="black") + labs(y="p-value", title="Trendline of Hosmer-Lemeshow Goodness of Fit Test") + geom_hline(yintercept = .05, col="red")


#Compare brier scores
bs_mod1 <- mean((pred_mod1 - ICUMort.f$in_hospital_death)^2)
bs_mod1
bs_mod2 <- mean((pred_mod2 - ICUMort.f$in_hospital_death)^2)
bs_mod2
bs_mod3 <- mean((pred_mod3 - ICUMort.f$in_hospital_death)^2)
bs_mod3
bc_null <- mean((pred_null - ICUMort.f$in_hospital_death)^2)
bc_null
#Almost the same, more parsimonious model acceptible - red_int_mod??

#Brier score indicates that the reduced and full models are better than the null, but nor by a huge degree.
#PseudoR2
library(DescTools)
#PsuedoR2 models under question
a <- t(PseudoR2(f.mod1, which = "all"))
b <- t(PseudoR2(f.mod2, which = "all"))
c <- t(PseudoR2(f.mod3, which = "all"))
d <- t(PseudoR2(null_mod, which = "all"))

#turn into table for easier interpretation
psudortbl <- rbind(a,b,c,d)
rownames(psudortbl) <- c("f.mod1", "f.mod2", "f.mod3", "null_mod")
#round values for output
round(psudortbl,4)
```

```{r AUROC for all models}


SAPS.mod <- glm(in_hospital_death ~ SAPS1, data = ICUMort.t, family = binomial)
predSAPS <- predict(SAPS.mod, type="response")

par(pty="s")
plot.roc(ICUMort.t$in_hospital_death, ICUMort.t$predprob, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(1,1,0.1, alpha = 0.3), col="yellow", legend= TRUE, legend.title="models")

#ICUMort.f %>% mutate(predprob=predict(red_int_mod, type="response")) -> ICUMort.f
plot.roc(ICUMort.f$in_hospital_death, pred_mod2, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.1,0.1,0.9, alpha = 0.3),add=TRUE, col="blue", legend= TRUE, legend.title="models")

#ICUMort.f %>% mutate(predprob=predict(sig_full_mod, type="response")) -> ICUMort.f
plot.roc(ICUMort.f$in_hospital_death, pred_mod3, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.1,0.9,0.1, alpha = 0.3),add=TRUE, col="green", legend= TRUE, legend.title="models")

plot.roc(ICUMort.f$in_hospital_death, pred_mod1, identity.col="black", legacy.axes=TRUE, add=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.9,0.1,0.1, alpha = 0.3), col="red", legend= TRUE, legend.title="models")

plot.roc(ICUMort.t$in_hospital_death, predSAPS, identity.col="black", legacy.axes=TRUE, add=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(0.1,1,1, alpha = 0.3), col="black", legend= TRUE, legend.title="models")

legend("bottomright",
       legend=c("Multivariable Model", "Interaction Model", "Multivariable Interation Model", "Combined Unit Specific Model", "SAPS Only Model"),
       title="Models:", 
       fill=c(rgb(.9,0.1,0.1, alpha = 0.3),rgb(.1,0.1,0.9, alpha = 0.3),rgb(.1,0.9,0.1, alpha = 0.3),rgb(1,1,0.1, alpha = 0.3),rgb(0.1,1,1, alpha = 0.3)), 
       border=c("red", "blue", "green", "yellow", "black"), 
       inset=c(0.1,0.1))
```


##Cath
##Selecting predictor variables for in hostital death

SAPS1 and SOFA seem like obvious choices, as these have been developed as predictors of clinical outcomes (SOFA) and risk of death (SAPS-I).  

Age should also be included, as those who are very young or very old will likely have a higher risk of death.

Include variables from SAPS-I as this has already been verified to correlate with increased risk of death (BUN, GCS, HCO3, HCT, HR, K, Na, RespRate, SysABP, Urine, WBC). Similarly for SOPS (PaO2, Platelets, Bilirubin, Creatinine, GCS).


```{r, Task 1 univariate models}
#Should start with univariate models for each of the variables we included.
#Create vector of variables
var_selection <- c("Glucose_diff", "WBC_min", "pH_diff", "Bilirubin_min", "BUN_min", "PaO2_max", "RespRate_min", "MAP_max", "Na_max", "GCS_max", "Age", "Gender", "ICUType","SAPS1")
#Create a list of univariate models for each of the variables selected
formulas<- paste('in_hospital_death ~', var_selection)
unimod_list <- lapply(formulas, glm, family = binomial(), data = ICUMort)
#Summary of all models
unimod_summ <- lapply(unimod_list, summary)
unimod_summ
#All variables except Na_max and sex are significant in univariate models
#Extract exp of coefficients - odds for each variable
unimod_odds <- lapply(unimod_list, function(x) {exp(coef(x))})
unimod_odds
#pH diff has the largest effect, others increase odds by between ~2 -15 %, most around 2.  Albumin_max decreases odds by ~ 38%.

#Check interactions from EDA
var_int_selection <- c("Glucose_diff:Gender", "BUN_min:Gender", "RespRate_min:Gender", "Bilirubin_min:Gender", "pH_diff:Gender", "BUN_min:ICUType", "PaO2_max:ICUType", "Bilirubin_min:ICUType", "GCS_max:ICUType", "Age:ICUType")
formulas_int<- paste('in_hospital_death ~', var_int_selection)
unimod_int_list <- lapply(formulas_int, glm, family = binomial(), data = ICUMort)
unimod_int_summ <- lapply(unimod_int_list, summary)
unimod_int_summ
#all interactions statistically significant.
```

```{r, multivariate models}
#Only include statisically significant variables from above. cut off 0.05.
full_mod<-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max + 
                Platelets_diff + RespRate_min, family = binomial(), data = ICUMort.f)
summary(full_mod)
#check statistically significant values
#create statistically significant model
sig_full_mod<-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max 
                , family = binomial(), data = ICUMort.f)
summary(sig_full_mod)

#compare full with statistically significant for anova and AIC
anova(full_mod, sig_full_mod, test = "Chisq")
full_mod$aic
sig_full_mod$aic
#test not significant, AIC lower for sig model, we can accept the simpler model

#check full models against stepwise removal of variables
red_full_mod <- step(full_mod, trace = 0)
anova(red_full_mod, sig_full_mod, test = "Chisq")
red_full_mod$aic
sig_full_mod$aic
#test not significant, AIC similar, we can accept the simpler model - sig_full_mod
f.mod1 <- sig_full_mod

#Statistically significant interactions from individual models above
int_mod<-glm(in_hospital_death ~  Age:Gender + BUN_min:Age + GCS_max:RespRate_min + GCS_max:Gender + Platelets_diff:Gender + RespRate_min:Gender + SAPS1:ICUType + Age:ICUType + Bilirubin_diff:ICUType + BUN_min:ICUType + GCS_max:ICUType + MAP_max:ICUType + PaO2_max:ICUType + Platelets_diff:ICUType + RespRate_min:ICUType + Gender:ICUType, family = binomial(), data = ICUMort.f)
summary(int_mod)
#check statistically significant values
# coef.df <- as.data.frame(summary(int_mod)$coefficients)
# as.matrix(coef.df[coef.df$`Pr(>|z|)`<0.05,])
#Age:Gender, BUN_min:Gender, pH_diff:Gender, BUN_min:ICUType, PaO2_max:ICUType, GCS_max:ICUType
#create statistically significant model
sig_int_mod<-glm(in_hospital_death ~ BUN_min:Age + GCS_max:Gender + SAPS1:ICUType + Bilirubin_diff:ICUType + BUN_min:ICUType + GCS_max:ICUType + PaO2_max:ICUType, family = binomial(), data = ICUMort.f)
summary(sig_int_mod)

#compare full with statistically significant for anova and AIC
anova(int_mod, sig_int_mod, test = "Chisq")
int_mod$aic
sig_int_mod$aic
#test statistically significant, AIC lower for int model, reject null hypothesis and keep original model

#check full models against stepwise removal of variables
red_int_mod <- step(int_mod, trace = 0)
anova(red_int_mod, int_mod, test = "Chisq")
red_int_mod$aic
int_mod$aic
#test not significant, AIC better for reduced model, we can accept the simpler model - red_int_mod
f.mod2 <- red_int_mod

#Full model with interactions
full_int_mod <-glm(in_hospital_death ~ SAPS1 + Age + Bilirubin_diff + BUN_min + GCS_max + ICUType + PaO2_max + 
                Platelets_diff + RespRate_min + Age:Gender + BUN_min:Age + GCS_max:RespRate_min + GCS_max:Gender +
                Platelets_diff:Gender + RespRate_min:Gender + SAPS1:ICUType + Age:ICUType + Bilirubin_diff:ICUType +
                BUN_min:ICUType + GCS_max:ICUType + MAP_max:ICUType + PaO2_max:ICUType + Platelets_diff:ICUType +
                RespRate_min:ICUType + Gender:ICUType, family = binomial(), data = ICUMort.f)
summary(full_int_mod)
#check statistically significant values
# coef.df <- as.data.frame(summary(full_int_mod)$coefficients)
# as.matrix(coef.df[coef.df$`Pr(>|z|)`<0.05,])
#pH_diff, BUN_min, Na_max, GCS_max, Age, ICUType, Gender, Age:Gender, PaO2_max:ICUType, GCS_max:ICUType

#create statistically significant model at 0.1
sig_full_int_mod<-glm(in_hospital_death ~ SAPS1 + GCS_max + ICUType + PaO2_max + Age:Gender + GCS_max:Gender +
                        SAPS1:ICUType + GCS_max:ICUType,
                      family = binomial(), data = ICUMort.f)
summary(sig_full_int_mod)

#compare full with statistically significant for anova and AIC
anova(full_int_mod, sig_full_int_mod, test = "Chisq")
full_int_mod$aic
sig_full_int_mod$aic
#test statistically significant, AIC lower for full int model, reject null hypothesis and keep original model

#check full models against stepwise removal of variables
red_full_int_mod <- step(full_int_mod, trace = 0)
anova(red_full_int_mod, full_int_mod, test = "Chisq")
red_full_int_mod$aic
full_int_mod$aic
#test not significant, AIC better for reduced model, we can accept the simpler model - red_full_int_mod
f.mod3 <- red_full_int_mod

#Compare chosen models to a null model
null_mod <- glm(in_hospital_death ~ 1, family = binomial, data = ICUMort.f)
#anova(sig_full_mod, null_mod, test = "Chisq")
anova(f.mod1, null_mod, test = "Chisq")
anova(f.mod2, null_mod, test = "Chisq")
anova(f.mod3, null_mod, test = "Chisq")
#All tests statistically significant - Reject null hypothesis of no relationship between predictors and outcome for all models

#compare chosen models
anova(f.mod1, f.mod2, test = "Chisq")
f.mod1$aic
f.mod2$aic
anova(f.mod1, f.mod3, test = "Chisq")
f.mod1$aic
f.mod3$aic
anova(f.mod2, f.mod3, test = "Chisq")
f.mod2$aic
f.mod3$aic
#all tests statistically significant, can not reject null hypothesis, AIC improved on reduced models, 
#unable to compare interaction models due to not enough degrees of freedom, 
#p-value more significant for reduced full interaction model and AIC lower that both other models


```
_will include some commentary on these results_

Next step is to examine the goodness of fit, and model assumptions.

```{r, Task1, goodness of fit}

#possible models - sig_full_mod, red_int_mod, red_full_int_mod

#Calculate the Hosmer-Lemeshow statistic
#linear predictors
library(ResourceSelection)
#Vector of predicted probabilities
pred_mod1 <- predict(sig_full_mod, type = "response")
hl_mod1 <- hoslem.test(sig_full_mod$y, pred_mod1, g=80)
hl_mod1
#No indication of a lack of fit as p > 0.05.

pred_mod2 <- predict(red_int_mod, type = "response")
hl_mod2 <- hoslem.test(red_int_mod$y, pred_mod2, g=80)
hl_mod2
#No indication of a lack of fit as p > 0.05.

pred_mod3 <- predict(red_full_int_mod, type = "response")
hl_mod3 <- hoslem.test(red_full_int_mod$y, pred_mod3, g=80)
hl_mod3
#No indication of a lack of fit as p > 0.05.

pred_null <- predict(null_mod, type = "response")


#Compare brier scores
(bs_mod1 <- mean((pred_mod1 - ICUMort.f$in_hospital_death)^2))
(bs_mod2 <- mean((pred_mod2 - ICUMort.f$in_hospital_death)^2))
(bs_mod3 <- mean((pred_mod3 - ICUMort.f$in_hospital_death)^2))
(bc_null <- mean((pred_null - ICUMort.f$in_hospital_death)^2))
#Almost the same, more parsimonious model acceptible - red_int_mod??

#Brier score indicates that the reduced and full models are better than the null, but nor by a huge degree.
#PseudoR2
library(DescTools)
#PsuedoR2 models under question
a <- t(PseudoR2(sig_full_mod, which = "all"))
b <- t(PseudoR2(red_int_mod, which = "all"))
c <- t(PseudoR2(red_full_int_mod, which = "all"))
d <- t(PseudoR2(null_mod, which = "all"))

#turn into table for easier interpretation
psudortbl <- rbind(a,b,c,d)
rownames(psudortbl) <- c("sig_full_mod", "red_int_mod", "red_full_int_mod", "null_mod")
#round values for output
round(psudortbl,4)

```

#Appears to be a strong difference between the physiological variables and what type of ICU they are in. This seems logical. Attempt creating a different model for individual ICUs and combining at the end for prediction outcome
```{r multimodel method for each ICU type}
#review analysis of ICUType specific models, create data frames
ICUMort.CCU <- ICUMort.t[ICUMort.t$ICUType=="Coronary Care Unit",]
ICUMort.CCU <- within(ICUMort.CCU, rm(ICUType))
ICUMort.MICU <- ICUMort.t[ICUMort.t$ICUType=="Medical ICU",]
ICUMort.MICU <- within(ICUMort.MICU, rm(ICUType))
ICUMort.SICU <- ICUMort.t[ICUMort.t$ICUType=="Surgical ICU",]
ICUMort.SICU <- within(ICUMort.SICU, rm(ICUType))
ICUMort.CSRU <- ICUMort.t[ICUMort.t$ICUType=="Cardiac Surgery Recovery Unit",]
ICUMort.CSRU <- within(ICUMort.CSRU, rm(ICUType))

#full models using most statistically significant output from each variable group with interaction between age and gender
CCU.full.mod <- glm(in_hospital_death ~ Age + Bilirubin_max + BUN_max + GCS_max +
                      Gender + Glucose_min + MAP_max + Na_diff + PaO2_max + pH_diff +
                      Platelets_diff + RespRate_min + TroponinI_min + WBC_max + SAPS1 +
                      Age:Gender
                      , data=ICUMort.CCU, family="binomial")
(sum.full.CCU <- summary(CCU.full.mod))
#significant predictors at 0.1 - BUN_max, GCS_max, SAPS1

MICU.full.mod <- glm(in_hospital_death ~ Age + Bilirubin_min + BUN_min + GCS_max +
                       Gender + Glucose_min + MAP_max + Na_max + PaO2_max + pH_min +
                       Platelets_min + RespRate_max + TroponinI_max + WBC_max + SAPS1 +
                       Age:Gender
                       , data=ICUMort.MICU, family="binomial")
(sum.full.MICU <- summary(MICU.full.mod))
#significant predictors at 0.1 - Age, BUN_min, Na_max, Platelets_dif, WBC_max, SAPS1


SICU.full.mod <- glm(in_hospital_death ~ Age + Bilirubin_diff + BUN_min + GCS_max + Gender +
                       Glucose_diff + MAP_diff + Na_min + PaO2_diff + pH_diff + Platelets_max +
                       RespRate_min + TroponinI_diff + WBC_min + SAPS1 + Age:Gender
                     , data=ICUMort.SICU, family="binomial")
(sum.full.SICU <- summary(SICU.full.mod))
#significant predictors at 0.05 - Age, BUN_min,GCS_max, Gender, pH_diff, RespRate_min, TroponinI_diff, WBC_min, SAPS1

CSRU.full.mod <- glm(in_hospital_death ~ Age + Bilirubin_max + BUN_max + GCS_max + Gender + 
                       Glucose_min + MAP_min + Na_diff + PaO2_min + pH_max + Platelets_max +
                       RespRate_max + TroponinI_diff + WBC_min + SAPS1 + Age:Gender
                     , data=ICUMort.CSRU, family="binomial")
(sum.full.CSRU <- summary(CSRU.full.mod))
#significant predictors at 0.05 - BUN_min, GCS_max, Na_diff, PaO2_max, pH_max, Platelets_max, WBC_min

#significan models - keeping Age:Gender as a predictor
#CCU
CCU.sig.mod <- glm(in_hospital_death ~ BUN_max + GCS_max + 
                      SAPS1 + Age:Gender, data=ICUMort.CCU, family="binomial")
(sum.sig.CCU <- summary(CCU.sig.mod))
anova(CCU.full.mod, CCU.sig.mod, test="Chisq")
CCU.full.mod$aic
CCU.sig.mod$aic
#not statistically different, AIC much better for simpler model
CCU.red.mod <- step(CCU.full.mod, trace=0)
#AIC stepwise reduced model included many 
(sum.red.CCU <- summary(CCU.red.mod))
anova(CCU.full.mod, CCU.red.mod, test="Chisq")
CCU.full.mod$aic
CCU.red.mod$aic
#significant and reduced models quite different as Age:Gender interaction kept. Unable to improve AIC without removing, neither are statistically different from the full model, AICs similar (dif <4)
#keep significant model.
CCU.mod <- CCU.sig.mod

#MICU
MICU.sig.mod <- glm(in_hospital_death ~ Bilirubin_min + BUN_min + Na_max + TroponinI_max +
                       WBC_max + SAPS1 + Age:Gender, data=ICUMort.MICU, family="binomial")
(sum.sig.MICU <- summary(MICU.sig.mod))
anova(MICU.full.mod, MICU.sig.mod, test="Chisq")
MICU.full.mod$aic
MICU.sig.mod$aic
#statistically significant difference, resid dev better in full model, AIC marginal. check stepwise reduction.
MICU.red.mod <- step(MICU.full.mod, trace=0)
anova(MICU.full.mod, MICU.red.mod, test="Chisq")
MICU.full.mod$aic
MICU.red.mod$aic
#not statistically different, AIC much improved in reduced model. Age:Gender interaction retained in reduced model - keep reduced
MICU.mod <- MICU.red.mod

#SICU
SICU.sig.mod <- glm(in_hospital_death ~ Age + BUN_min + GCS_max + Glucose_diff + PaO2_diff + pH_diff + 
                       Platelets_max + RespRate_min + WBC_min + SAPS1 + Age:Gender, data=ICUMort.SICU, family="binomial")
(sum.sig.SICU <- summary(SICU.sig.mod))
anova(SICU.full.mod, SICU.sig.mod, test="Chisq")
SICU.full.mod$aic
SICU.sig.mod$aic
#no statistically significant difference, AIC slightly better for sig model. check stepwise reduction.
SICU.red.mod <- step(SICU.full.mod, trace=0)
anova(SICU.full.mod, SICU.red.mod, test="Chisq")
SICU.full.mod$aic
SICU.red.mod$aic
#not statistically different, AIC slightly improved in reduced model. Age and Gender in reduced model without interaction, keep reduced model
SICU.mod <- SICU.red.mod


CSRU.sig.mod <- glm(in_hospital_death ~ BUN_max + GCS_max + MAP_min + Na_diff + Platelets_max +
                      pH_max + SAPS1 + Age:Gender, data=ICUMort.CSRU, family="binomial")
(sum.sig.CSRU <- summary(CSRU.sig.mod))
anova(CSRU.full.mod, CSRU.sig.mod, test="Chisq")
CSRU.full.mod$aic
CSRU.sig.mod$aic
#not statistically different, AIC improved in sig model. Added platelets from stepwise reduction with good effect
CSRU.red.mod <- step(CSRU.full.mod, trace=0)
anova(CSRU.full.mod, CSRU.red.mod, test="Chisq")
CSRU.full.mod$aic
CSRU.red.mod$aic
#not statistically different, AIC improved in reduced model. AIC difference <1 between sig and red mod. Keep sig model.
CSRU.mod <- CSRU.sig.mod


#check goodness of fit
#Calculate the Hosmer-Lemeshow statistic
#linear predictors
library(ResourceSelection)
#Vector of predicted probabilities
pred.CCU <- predict(CCU.mod, type = "response")
(hl_CCU <- hoslem.test(CCU.mod$y, pred.CCU, g=30))
#No indication of a lack of fit as p > 0.05.

pred.MICU <- predict(MICU.mod, type = "response")
(hl.MICU <- hoslem.test(MICU.mod$y, pred.MICU, g=30))
#No indication of a lack of fit as p > 0.05.

pred.SICU <- predict(SICU.mod, type = "response")
(hl.SICU <- hoslem.test(SICU.mod$y, pred.SICU, g=30))
#No indication of a lack of fit as p > 0.05.

pred.CSRU.sig <- predict(CSRU.sig.mod, type = "response")
(hl.CSRU <- hoslem.test(CSRU.sig.mod$y, pred.CSRU.sig, g=40))
#h-l statistically significant, suggests lack of fit. Try reduced model instead.
pred.CSRU.red <- predict(CSRU.red.mod, type = "response")
(hl.CSRU <- hoslem.test(CSRU.red.mod$y, pred.CSRU.red, g=40))
#h-l statistically significant, suggests lack of fit. Try full model instead.
pred.CSRU <- predict(CSRU.full.mod, type = "response")
(hl.CSRU <- hoslem.test(CSRU.full.mod$y, pred.CSRU, g=40))
#better fit
CSRU.mod <- CSRU.full.mod

null.CCU.mod <- glm(in_hospital_death ~ 1, data = ICUMort.CCU, family = binomial)
null.MICU.mod <- glm(in_hospital_death ~ 1, data = ICUMort.MICU, family = binomial)
null.SICU.mod <- glm(in_hospital_death ~ 1, data = ICUMort.SICU, family = binomial)
null.CSRU.mod <- glm(in_hospital_death ~ 1, data = ICUMort.CSRU, family = binomial)

null.CCU <- predict(null.CCU.mod, type = "response")
null.MICU <- predict(null.MICU.mod, type = "response")
null.SICU <- predict(null.SICU.mod, type = "response")
null.CSRU <- predict(null.CSRU.mod, type = "response")

#Compare brier scores
(bs.CCU <- mean((pred.CCU - ICUMort.CCU$in_hospital_death)^2))
mean((null.CCU - ICUMort.CCU$in_hospital_death)^2)

(bs.MICU <- mean((pred.MICU - ICUMort.MICU$in_hospital_death)^2))
mean((null.MICU - ICUMort.MICU$in_hospital_death)^2)

(bs.SICU <- mean((pred.SICU - ICUMort.SICU$in_hospital_death)^2))
mean((null.SICU - ICUMort.SICU$in_hospital_death)^2)

(bs.CSRU <- mean((pred.CSRU - ICUMort.CSRU$in_hospital_death)^2))
mean((null.CSRU - ICUMort.CSRU$in_hospital_death)^2)

#brier scores better than null models

#Brier score indicates that the reduced and full models are better than the null, but nor by a huge degree.
#PseudoR2
library(DescTools)
#PsuedoR2 models under question
a <- t(PseudoR2(CCU.mod, which = "all"))
b <- t(PseudoR2(MICU.mod, which = "all"))
c <- t(PseudoR2(SICU.mod, which = "all"))
d <- t(PseudoR2(CSRU.mod, which = "all"))
e <- t(PseudoR2(null_mod, which = "all"))

#turn into table for easier interpretation
psudortbl <- rbind(a,b,c,d,e)
rownames(psudortbl) <- c("CCU.mod", "MICU.mod", "SICU.mod", "CSRU.mod","null_mod")
#round values for output
round(psudortbl,4)

```


```{r multi unit analysis prediction}

ICUMort.t[ICUMort.t$ICUType=="Coronary Care Unit",]

ICUMort.t[ICUMort.t$ICUType=="Medical ICU",]

ICUMort.t[ICUMort.t$ICUType=="Surgical ICU",]

ICUMort.t[ICUMort.t$ICUType=="Cardiac Surgery Recovery Unit",]

ICUMort.t$predprob <- 0

ICUMort.t[ICUMort.t$ICUType=="Coronary Care Unit",] %>% mutate(predprob=predict(CCU.mod, type="response")) -> ICUMort.t[ICUMort.t$ICUType=="Coronary Care Unit",]

ICUMort.t[ICUMort.t$ICUType=="Medical ICU",] %>% mutate(predprob=predict(MICU.mod, type="response")) -> ICUMort.t[ICUMort.t$ICUType=="Medical ICU",]

ICUMort.t[ICUMort.t$ICUType=="Surgical ICU",] %>% mutate(predprob=predict(SICU.mod, type="response")) -> ICUMort.t[ICUMort.t$ICUType=="Surgical ICU",]

ICUMort.t[ICUMort.t$ICUType=="Cardiac Surgery Recovery Unit",] %>% mutate(predprob=predict(CSRU.mod, type="response")) -> ICUMort.t[ICUMort.t$ICUType=="Cardiac Surgery Recovery Unit",]

ICUMort.t


par(pty="s")
plot.roc(ICUMort.t$in_hospital_death, ICUMort.t$predprob, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.9,0.1,0.1, alpha = 0.3), col="red", legend= TRUE, legend.title="models")
```

```{r AUROC for all models}



```

##Goodness of fit
Both the full and reduced model explain more variation than the null model, but it is a relatively small amount of the variance (McFadden ~0.17).  No evidence of a lack of fit from the Hosmer-Lemeshow statistic, and Brier scores are lower (i.e. better) for both full ad reduced model compared to the null.  Not a huge difference between the full and reduced models.

```{r}
#Calculate confusion matrix
library(caret)
#Create vector of predicted outcomes, set 0.5 as threshold
pred_oc <- as.factor(ifelse(pred < 0.5, "no", "yes"))
#Convert outcome to no yes factor
oc <- as.factor(ifelse(ICUMort$in_hospital_death == 0, "no", "yes"))
confusionMatrix(pred_oc, oc, positive = "yes")
#Not a great at predicting death, only ~16% sensitivity.  Look at distribution of predicted probs for each case.
mod_pred <- data.frame(pred,oc)
ggplot(mod_pred, aes(x = oc, y = pred)) + geom_boxplot() +
  labs(title = "Predicted probabilities from reduced model by outcome", 
       x = "Death", y = "Probability")
#Based on the graphs, change the threshold for considering the outcome to occur to 0.2
#Create vector of predicted outcomes, set 0.5 as threshold
pred_oc2 <- as.factor(ifelse(pred < 0.2, "no", "yes"))
#Convert outcome to no yes factor
confusionMatrix(pred_oc2, oc, positive = "yes")
#Still not great at predicting the outcome, sensitivity improved, but specificity is reduced - pos predictive value only 35%. Should look to see how this compares to 
#First try ROC
library(Epi)
#Plot reduced model ROC
ROC(pred, oc, plot = "ROC", main = "Reduced interaction model ROC")
#Plot full model ROC
pred_full <- predict(full_mod, type = "response")
ROC(pred_full, oc, plot = "ROC", main = "Full model ROC")

```

##Predictive value
Moderate performance of the model, with ROC AUC of 0.786.  By comparison, the mean SOFA score showed AUC of ~0.88 in the publication linked above.  

Maybe if more variables are included from the initial set, a better prediction could be achieved?

What about SOFA alone?
```{r - check AUC of univariate SOFA model}
ROC(predict(unimod_list[[1]], type = "response"), oc, plot = "ROC")
#AUC only 0.651 in this dataset for a univariate SOFA model.

#all AUROCs for current models (including one for SAPS only)
SAPS.mod <- glm(in_hospital_death ~ SAPS1, data = ICUMort.t, family = binomial)
predSAPS <- predict(SAPS.mod, type="response")

par(pty="s")
plot.roc(ICUMort.t$in_hospital_death, ICUMort.t$predprob, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(1,1,0.1, alpha = 0.3), col="yellow", legend= TRUE, legend.title="models")

#ICUMort.f %>% mutate(predprob=predict(red_int_mod, type="response")) -> ICUMort.f
plot.roc(ICUMort.f$in_hospital_death, pred_mod2, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.1,0.1,0.9, alpha = 0.3),add=TRUE, col="blue", legend= TRUE, legend.title="models")

#ICUMort.f %>% mutate(predprob=predict(sig_full_mod, type="response")) -> ICUMort.f
plot.roc(ICUMort.f$in_hospital_death, pred_mod3, identity.col="black", legacy.axes=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.1,0.9,0.1, alpha = 0.3),add=TRUE, col="green", legend= TRUE, legend.title="models")

plot.roc(ICUMort.f$in_hospital_death, pred_mod1, identity.col="black", legacy.axes=TRUE, add=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(.9,0.1,0.1, alpha = 0.3), col="red", legend= TRUE, legend.title="models")

plot.roc(ICUMort.t$in_hospital_death, predSAPS, identity.col="black", legacy.axes=TRUE, add=TRUE,
    auc.polygon = TRUE, auc.polygon.col=rgb(0.1,1,1, alpha = 0.3), col="black", legend= TRUE, legend.title="models")

legend("bottomright",
       legend=c("Multivariable Model", "Interaction Model", "Multivariable Interation Model", "Combined Unit Specific Model", "SAPS Only Model"),
       title="Models:", 
       fill=c(rgb(.9,0.1,0.1, alpha = 0.3),rgb(.1,0.1,0.9, alpha = 0.3),rgb(.1,0.9,0.1, alpha = 0.3),rgb(1,1,0.1, alpha = 0.3),rgb(0.1,1,1, alpha = 0.3)), 
       border=c("red", "blue", "green", "yellow", "black"), 
       inset=c(0.1,0.1))
```

I wonder if we should try to include SAPS1 with missing rows removed from the entire dataset - would mean loosing 96 of 2061.  Could also try adding other variables ruled out above to the model to see if we get a better fit.






# Task 2 (15 marks)

In this task, you are required to develop a Cox proportional hazards survival model using the `icu_patients_df1` data set which adequately explains or predicts the length of survival indicated by the `Days` variable, with censoring as indicated by the `Status` variable.  You should fit a series of models, maybe three or four, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge. Aim for between five and ten predictor variables (slightly more or fewer is OK). It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct, but do you best). You should assess each model you consider for goodness of fit and other relevant statistics, and you should assess your final model for violations of assumptions and perform other diagnostics which you think are relevant (and modify the model if indicated, or at comment on the possible impact of what your diagnostics show). Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find.


### Hints

1. Select an initial subset of explanatory variables that you will use to predict survival. Justify your choice.

2. Conduct basic exploratory data analysis on your variables of choice.

3. Fit appropriate univariate Cox proportional hazards models.

4. Fit an appropriate series of multivariable Cox proportional hazards models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

5. Present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge.

6. For your final model, present a set of diagnostic statistics and/or charts and comment on them. 

7. Write a very brief paragraph summarising the most important findings of your final model. Include the most important values from the statistical output, and a simple clinical interpretation. 

####
**Create your response to this task here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text.** 

```{r data-setup}
library(survival)
```

### Task 2 Part1.a
```{r Taks2_Part1a}
# load the data
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
#View(icu_patients_df1)
head(icu_patients_df1)
str(icu_patients_df1)

```




# KD code

##Consideration for variables
We only need to include 5-10ish variables. This is requires a logical approach to culling. Propose first step is to remove any variables that do not have a full set of data (eg 2061 records). The rationale for this is that if you include variables incomplete variables then if you drop an incomplete variable you may not be able to adequately compare the models. Suggest 1 variable per body system not covered in SOFA score (GCS, MAP, Creatinine/Urine, PaO2/FiO2, Bilirubin, Platelets) as SOFA covers those variables.  
  
# Consider using variables used in already validated scores such as SAPSII, APACHE etc. https://www.mdcalc.com/simplified-acute-physiology-score-saps-ii
# Also we want to use markers which will predict the length of survival. Often a change in these markers is a bad sign - so consider using diff variables more. 

#Variables left after incomplete variables removed and variables in SOFA score.

##Assessable outcome: "in_hospital_death"  
_Demographics: "Age", "Gender", "ICUType"_
_Age and sex are demographics important for identifying confounders. Height is missing a lot of values. Is length of stay a predictor or an outcome? Suggest outcome, remove.  Weight can be bad for overweight and underweight._
##Selected Variables: "Age", "Gender", "ICUType" 
Total number of variables = 3

_CNS: "Na_diff", "Na_max", "Na_min"_
_Low sodium worse than high sodium but both bad.   _
##Selected Variables: "Na_diff"
Total number of variables = 4

_Cardiovascular: "HR_diff", "HR_max", "HR_min", "K_diff", "K_max", "K_min", "Mg_diff", "Mg_max", "Mg_min", "Temp_diff", "Temp_max", "Temp_min", "TroponinI_diff", "TroponinI_max", "TroponinI_min", "TroponinT_diff", "TroponinT_max", "TroponinT_min"_
_Troponins are a sign of cardiac ischaemia, which can occur for multiple reasons and would be a poor prognostic indicator. Trop I and T are similar, use Trop I. High temp is a sign of sepsis/infection. K is an electrolyte that often becomes deranged, both high and low are equally bad - use K_diff. HR is bad at both extremes - use HR_diff._
##Selected Variables: "HR_diff", "K_diff", "Temp_max", "TroponinI_max"
Total number of variables = 8

_PULMONARY: "PaCO2_diff", "PaCO2_max", "PaCO2_min", "RespRate_diff", "RespRate_max", "RespRate_min", "SaO2_diff", "SaO2_max", "SaO2_min"_
_These are covered by the SOFA score, which takes into account the PaO2 and whether they are being mechanically ventilated. Leave all of them out.._
##Selected Variables: none
Total number of variables = 8

_Renal: "BUN_diff", "BUN_max", "BUN_min"_
_Good marker of renal impairment, higher is worse._
##Selected Variables: "BUN_max"
Total number of variables = 9

_Liver: "Albumin_diff", "Albumin_max", "Albumin_min", "ALP_diff", "ALP_max", "ALP_min", "ALT_diff", "ALT_max", "ALT_min", "AST_diff", "AST_max", "AST_min", "Cholesterol_diff", "Cholesterol_max", "Cholesterol_min"_
_Low albumin is bad and is a marker of the liver's synthetic ability, however it is covered by the bilirubin measurement. High bilirubin is bad - used in the SOFA score. ALT and ALT are early markers for liver failure and may be an early predictor for mortality - try use ALT_max, (ALT better than AST for most liver issues). Unclear about the role of Cholesterol in critical illness, leave out._
##Selected Variables: "ALT_max"
Total number of variables = 10

_METABOLIC/OTHER: "Glucose_diff", "Glucose_max", "Glucose_min", "HCO3_diff", "HCO3_max", "HCO3_min", "Lactate_diff", "Lactate_max", "Lactate_min", "WBC_diff", "WBC_max", "WBC_min", "pH_diff", "pH_max", "pH_min", "HCT_diff", "HCT_max", "HCT_min"  _
_Glucose is often high at times of acute illness and can be controlled with insulin. Glucose in this setting is probably a poor indicator of comorbid diabetes, leave these out. HCO3 bad if it’s low - use this. High lactate is bad, try use this. WBC marker for infection in both directions, but low <1 is really bad - use this. Low platelets are bad. pH is a good marker of severity of illness, bad in both directions, also related to HCO3 and PaCO2. HCT used in APACHE scoring, low is bad, add this in as well._
##Selected Variables: "HCO3_min", "Lactate_max", "WBC_min", "pH_diff", "HCT_min"
Total number of variables = 15

_SCORES: "SOFA"_
_SOFA score uses GCS, MAP, Creatinine/Urine, PaO2/FiO2, Bilirubin and Platelets._
##Selected Variables: "SOFA" 
Total number of variables = 16

Final list of variables:
"SOFA", "Age", "Gender", "ICUType", "Na_diff", "HR_diff", "K_diff", "Temp_max", "TroponinI_max", "BUN_max", "ALT_max", "HCO3_min", "Lactate_max", "WBC_min", "pH_diff", "HCT_min"

```{r}
#remove incomplete variables to maintain data integrity
ICUsurv <- icu_patients_df1[,complete.cases(t(icu_patients_df1))]

#ensure all max and min values are in the right columns

#function to swap values that are inconsistent with allocation
maxval <- function(df, Max, Min){
  df[na.omit(df[,Max])<na.omit(df[,Min]),c(Min,Max)] <- unname(df[na.omit(df[,Max])<na.omit(df[,Min]),c(Max,Min)])
  df
}

#loop checks for "max" variables then compares to the corresponding "min" value and swaps values where necessary
for (i in names(ICUsurv)){
  if (substr(i, nchar(i)-2, nchar(i)) == "max"){
  j <- paste0(substr(i, 1, nchar(i)-4 ),"_min")
  t.ICUsurv <- maxval(ICUsurv, i, j)
  }
}

#dataframe with selected varialbles only
ICUsurv.f <- ICUsurv[,c("SOFA", "Age", "Gender", "ICUType", "Na_diff", "HR_diff", "K_diff", "Temp_max", "TroponinI_max", "BUN_max", "ALT_max", "HCO3_min", "Lactate_max", "WBC_min", "pH_diff", "HCT_min", "in_hospital_death", "Status", "Days")]

require("ggplot2")
library(ggplot2)

for (i in names(ICUsurv.f)){
  if(is.numeric(ICUsurv.f[[i]])){
  
    print(ggplot(ICUsurv.f, aes(ICUsurv.f[[i]])) +
          geom_histogram(aes(y = ..density.., fill = factor(in_hospital_death, labels = c("No","Yes"))), alpha = .3, bins = 30, position = "identity") +
          labs(x=i, fill="Died", title = paste0("Density Plot of ", i,", ", length(na.omit(ICUsurv.f[[i]]))," values available")))

    }else{

    print(ggplot(ICUsurv.f, aes(ICUsurv.f[[i]], fill=factor(in_hospital_death, labels = c("No", "Yes")))) +
          geom_bar(position = "fill") +
          labs(x = i, fill = "Died", title = paste0("Bar Plot of ", i,", ",  length(na.omit(ICUsurv.f[[i]])), " values available")))
  
  }

}


```

```{r 2c}




vars_2 <- c("SOFA", "Age", "Gender", "ICUType", "Na_diff", "HR_diff", "K_diff", "Temp_max", "TroponinI_max", "BUN_max", "ALT_max", "HCO3_min", "Lactate_max", "WBC_min", "pH_diff", "HCT_min")

#making formulas
univ_formulas <- sapply(vars_2, function(x)as.formula(paste('Surv(Days,Status)~',x))
)
#making a list of models
univ_models <- lapply(univ_formulas, function(x){coxph(x,data=ICUsurv.f)})

#Create a list of univariate Cox PH models for each of the variables selected
#formulas2<- paste0("Surv(Days, Status) ~ ", vars_2)
#unicox_list <- lapply(formulas2, coxph, data = t.ICUsurv)
#summary(unicox_list)


univ_results <- lapply(univ_models, function(x){summary(x)})
univ_results
```
TroponinI_max is clearly not significant - remove.
HCT_min also not significant. Also this is related to pH - remove. 
##HCT not related to pH, HCO3 and pH are related
Leave others in for now. 

```{r}

cox.multi <- coxph(Surv(Days,Status) ~ SOFA + Age + Gender + ICUType + Na_diff + HR_diff + K_diff + Temp_max + BUN_max + ALT_max + Lactate_max + WBC_min + pH_diff + HCT_min, data=ICUsurv.f)

summary(cox.multi)



```

```{r}

cox.s1 <- step(cox.multi)
summary(cox.s1)



```






# Yalin's code #

### Task 2 Part1.b
The categorical variables in the dataset that may be associated with a person’s survival are:
in_hospital_death;
ICUType;
Gender.

The continuous variables in the dataset that may be associated with a person’s survival are:
Length_of_stay;
SAPS1: the risk of death in ICU patients;
SOFA: trends in organ dysfunction;
Survival;
Age;
all 3 summary variables for 36 clinical measures.

### Task 2 Part2
```{r Taks2_Part2.a}
# Basic exploratory data analysis for categorical variables:

cat("Frequency for the categories of in_hospital_death variable")
# use table function to calculate frequency
ihd_table <- table(icu_patients_df1$in_hospital_death)
# use addmargins function to add total to the table
addmargins(ihd_table)
# calculate proportions from the frequency table
# use round function to keep 2 decimal digit
round(prop.table(ihd_table), digits=2)
# As shown in the result: 0(survivor) = 86%; 1(died in-hospital) = 14% 

cat("Frequency for the categories of ICUType variable")
type_table <- table(icu_patients_df1$ICUType)
addmargins(type_table)
round(prop.table(type_table), digits=2)
# As shown in the result: Coronary Care Unit=14%; Surgery Recovery Unit=22%; Medical ICU=38%; Surgical ICU=26%

cat("Frequency for the categories of Gender variable")
Gender_table <- table(icu_patients_df1$Gender)
addmargins(Gender_table)
round(prop.table(Gender_table), digits=2)
# As shown in the result: Female=44%; Male=56%

```


```{r Taks2_Part2.b}
# select only the rows with death as an outcome
hospitaldeath <- icu_patients_df1[icu_patients_df1$in_hospital_death == 1, ]

# Plot the overall Kaplan-Meier survival curve for the hospitaldeath data
hd_fit <- survfit( Surv(Days,Status) ~ 1, data = hospitaldeath) 
summary(hd_fit)

# plotting Kaplan-Meier estimate
plot(hd_fit, main = 'Kaplan-Meier estimate of survival function', xlab = 'Days')

# The Kaplan-Meier plot allows us to see how the survival function changes over time. 
# The dotted lines around the survival curve represents the 95% confidence interval.

# get the mean and median survival times
print(hd_fit, print.rmean = TRUE)
# plot the median survival time
plot(hd_fit, main = 'Kaplan-Meier estimate of survival function', xlab = 'Days')
abline(a = 0.5, b = 0)

# As shown in the table, there were 297 events (deaths) during the follow-up, and the median follow-up time until death is 8 days.

# plotting Nelson-Aalen estimate
plot(hd_fit, fun="cumhaz", main = "Nelson-Aalen estimate of cumulative hazard function", xlab = "Days")
# As shown in the plot, the confidence interval becomes wider after 50 days. 

# plot the Kaplan-Meier survival curves by the categorical variables
par(mfrow=c(2,2))
hd_fitbyihd <- survfit( Surv(Days,Status) ~ in_hospital_death, data = hospitaldeath) 
summary(hd_fitbyihd)
plot(hd_fitbyihd, main = 'Kaplan-Meier estimate of survival function', xlab = 'in-hospital death',col = c("red","blue"))
# 0:survivor; 1:died in-hospital

hd_fitbyGender <- survfit( Surv(Days,Status) ~ Gender, data = hospitaldeath) 
summary(hd_fitbyGender)
plot(hd_fitbyGender, main = 'Kaplan-Meier estimate of survival function', xlab = 'in-hospital death',col = c("red","blue"))

hd_fitbytype <- survfit( Surv(Days,Status) ~ ICUType, data = hospitaldeath) 
summary(hd_fitbytype)
plot(hd_fitbytype, main = 'Kaplan-Meier estimate of survival function', xlab = 'ICU Type',col = c("red","blue","orange","black"))

# calculate log-rank tests for differences in survival experience by the categorical variable
# get the median survival times by in_hospital_death
print(hd_fitbytype) 
# 2-sample log-rank test
hd_fitbytype2 <- coxph( Surv(Days,Status) ~ ICUType, data = hospitaldeath) 
summary(hd_fitbytype2)

print(hd_fitbyGender) 
# 2-sample log-rank test
hd_fitbyGender2 <- coxph( Surv(Days,Status) ~ Gender, data = hospitaldeath) 
summary(hd_fitbyGender2)
```
The non-significant log-rank test of equality of the survival function means that the null hypothesis cannot be rejected, and there is no significant difference between groups. The variables with non-significant log-rank test are ICUType (p-value = 0.4) and Gender (p-value = 0.8).


```{r Taks2_Part2.c}
# remove observations that contain any missing values
fitbyall <- coxph( Surv(Days,Status) ~ . , data = icu_patients_df1, na.action=na.omit) 
summary(fitbyall)

# sort p-value 
fitallp <- data.frame(sort(summary(fitbyall)$coef[summary(fitbyall)$coef[,5] <= .05, 5], decreasing = FALSE))
names(fitallp) <- "Pr(>|z|)"
fitallp

# 11 most significant variables: 
# ALP_max, ALP_min, Cholesterol_min, Creatinine_max, Creatinine_min
# MAP_diff, MAP_max, NIMAP_min, pH_min, Urine_diff, Urine_max

```


```{r Taks2_Part2.d}
# Basic exploratory data analysis on variables of choice

# ALP_max, ALP_min, Cholesterol_min, Creatinine_max, Creatinine_min
# MAP_diff, MAP_max, NIMAP_min, pH_min, Urine_diff, Urine_max

summary(icu_patients_df1[,c("ALP_max","ALP_min","Cholesterol_min","Creatinine_max","Creatinine_min",
                            "MAP_diff","MAP_max","NIMAP_min","pH_min","Urine_diff", "Urine_max")])

```

### Task 2 Part3
```{r Taks2_Part3}
# fit Cox PH Model to significant predictor variables
vars_3 <- c("ALP_max", "ALP_min", "Cholesterol_min", "Creatinine_max", "Creatinine_min", "MAP_diff", "MAP_max", "NIMAP_min", "pH_min", "Urine_diff", "Urine_max")

univ_formulas <- sapply(vars_3, function(x)as.formula(paste('Surv(Days,Status)~',x)))
univ_models <- lapply(univ_formulas, function(x){coxph(x,data=icu_patients_df1)})

univ_results <- lapply(univ_models, function(x){summary(x)})
univ_results
```

The maximum likelihood estimate of the regression coefficient (coef) for ALP_max is 0.0012183 (a positive sign means that the hazard (risk of death) is higher), and thus hazard ratio for ALP_max is exp(0.0012183) = 1.0012190.  
The wald statistic evaluates, whether the beta coefficient of a given variable is statistically significantly different from 0. From the output above, we can conclude that the variable ALP_max have statistically significant coefficients (p-value=0.000399).
The output also provides a 95% confidence interval for ALP_max, 1.001 to 1.002.  
Finally, the output gives p-values for three alternative tests for overall significance of the model: the likelihood-ratio test, Wald test, and score logrank statistics. As shown in the result, the p-value for the overall significance of the model is less than 0.05. 


Non-significant: MAP_max



### Task 2 Part4
```{r Taks2_Part4}
# fit multivariable Cox proportional hazards models with significant variables above
cox.mv1 <- coxph(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  +  MAP_max + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)
summary(cox.mv1)

# Not significant in multivariable model
# ALP_max, ALP_min, Creatinine_max, Creatinine_min, MAP_diff, MAP_max

# fit multivariable Cox proportional hazards models with significant univariate Cox proportional hazards models
cox.mv2 <- coxph(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)
summary(cox.mv2)

# Not significant in multivariable model
# ALP_max, ALP_min, Creatinine_max, Creatinine_min

# The p-value (<2e-16) for all three overall tests (likelihood ratio test, Wald test, and score test) are significant, indicating that the model is significant.

cox.mv3 <- coxph(Surv(Days, Status) ~ Cholesterol_min + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)
summary(cox.mv3)

print(anova(cox.mv1, cox.mv2))
print(anova(cox.mv2, cox.mv3))


# If the p-value is not sufficiently low (usually greater than 0.05), we should favor the simpler model
# As shown in the result above, the p-value (0.7971) is not significant, cox.mv2 will be chosen.
#  p-value (0.009266) is significant, cox.mv2 will be chosen.


# the AICs for the two models
extractAIC(cox.mv1)
extractAIC(cox.mv2)
extractAIC(cox.mv3)
extractAIC(cox.multi)

# Higher (worse) AIC for the cox.mv3 model


```


### Task 2 Part5
```{r Taks2_Part5}
# final model:
cox.mv2 <- coxph(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)
summary(cox.mv2)

```

### Task 2 Part6


Using semi-parametric methods:
advantage of Cox Proportional Hazards regression:
fit survival models without assuming the distribution


```{r Taks2_Part6.a}
# diagnostic statistics
# graphical diagnostic

# testing for proportional hazards assumption using cox.zph()
prop.mv1 <- cox.zph(cox.mv2)
prop.mv1
# scaled Schoenfeld residuals against the transformed time

library("survminer")
ggcoxzph(prop.mv1)

# interpretation of the graph in its most simplest way is how curved it is: 
# if it is fairly flat, the assumption of proportionality is not (much) violated

```

The output from the proportional hazards test shows that the test is statistically significant (p-value < 0.05) for ALP_max, ALP_min, Urine_diff, and Urine_max variable. The global test is also statistically significant (p-value = 1.52e-04). 


ALP_max, ALP_min, Urine_diff, and Urine_max variable:
violates the PH assumption, resulting in strong evidence of non-proportional hazards for the global model






```{r}
# assume that continuous covariates have a linear form

# plot the Martingale residuals against continuous covariates can be used to detect nonlinearity 
# graphs of continuous covariates against martingale residuals of null cox proportional hazards model
ggcoxfunctional(Surv(Days, Status) ~ ALP_max + log(ALP_max) + sqrt(ALP_max), data = icu_patients_df1)
ggcoxfunctional(Surv(Days, Status) ~ ALP_max + log(ALP_min) + sqrt(ALP_min), data = icu_patients_df1)
ggcoxfunctional(Surv(Days, Status) ~ Urine_diff + log(Urine_diff) + sqrt(Urine_diff), data = icu_patients_df1)


# Patterns of Urine_diff in the plot may suggest that the variable is not properly fit.

```




```{r}
# relative importance of covariates:
# rank the covariate values and plot relative hazard against ranks

# relative hazard is scaled to the reference hazard
# reference hazard is related to the median of covariates
# relative hazard can vary depending on the distribution of covariate values in a population


library("rankhazard")
cox.mv2 <- coxph(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit, x=TRUE)

rankhazardplot(cox.mv2, data= icu_patients_df1)



# horizontal axis: the minimum, first quartile, median, third quartile and maximum values of each covariate are reported

```


```{r}
# Hazard ratio curves for continuous predictors:
# Cox proportional hazards model assumes that the continuous predictors are in linear association with log-hazard

# relationship between continuous predictor and hazard ratio helps investigators to check for the linearity assumption

library("splines")
library("smoothHR")

cox.mv2 <- coxph(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit, x=TRUE)

hrplot <- smoothHR(data = icu_patients_df1, coxfit = cox.mv2)

plot(hrplot, predictor= "Creatinine_max", prob = 0, conf.level = 0.95)

```


```{r}

# plots the estimated changes in the regression coefficients upon deleting each observation in turn
ggcoxdiagnostics(cox.mv2, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())


ggcoxdiagnostics(cox.mv2, type = "deviance",
                 linear.predictions = FALSE, ggtheme = theme_bw())


```



```{r}
# Parametric methods:


w.mv1 <- eha::phreg(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)
plot(w.mv1)

pch800.mv1 <- eha::phreg(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit, dist = "pch", cuts = c(800))

pch1800.mv1 <- eha::phreg(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit, dist = "pch", cuts = c(1800))



mll.mv1 <- c(w.mv1$loglik[2], pch800.mv1$loglik[2], pch1800.mv1$loglik[2])
names(mll.mv1) <- c("w", "pch800", "pch1800")
mll.mv1

# The largest value is the best. 

cox.mv2 <- eha::coxreg(Surv(Days, Status) ~ ALP_max + ALP_min + Cholesterol_min + Creatinine_max + Creatinine_min + MAP_diff  + NIMAP_min + pH_min + Urine_diff + Urine_max, data = icu_patients_df1, na.action=na.omit)

# check graphical goodness-of-fit test
eha::check.dist(cox.mv2, w.mv1)
eha::check.dist(cox.mv2, pch800.mv1)
eha::check.dist(cox.mv2, pch1800.mv1)


```


### Task 2 Part7
summarising the most important findings of final model:

















## Save, knit and submit

**Reminder**: don't forget to save this file, to knit it to check that everything works, and then submit via the drop box in Openlearning.

## Submit your assignment

When you have finished, and are satisfied with your assignment solutions, and this file knits without errors and the output looks the way you want, then you should submit via the drop box in Openlearning.

### Problems?

If you encounter problems with any part of the process described above, please contact the course instructor via OpenLearning as soon as possible so that the issues can be resolved in good time, and well before the assignment is due.


### Additional Information

Each task attracts the indicated number of marks (out of a total of 30 marks for the assignment). The instructions are deliberately open-ended and less prescriptive than the individual assignments to allow you some latitude in what you do and how you go about the task. However, to complete the tasks and gain full marks, you only need to replicate or repeat the steps covered in the course - if you do most or all of the things described in the revalant chapters of the HDAT9600 course, full marks wil be awarded. 

Note also that with respect to the model fitting, there are no **right** or **wrong** answers when it comes to variable selection and other aspects of model specification. Deep understanding of the underlying medical concepts which govern patient treatment and outcomes in ICUs is not required or assumed, although you should try to gain some understanding of each variable using the links provided. You will not be marked down if your medical justifications are not exactly correct or complete, but do you best, and don't hesitate to seek help from the course instructor.
